On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-14 manuscripts/PNAS/chazthm_pnas.tex
refs fixed and minor typos in manu

\keywords{Sparse coding, dictionary learning, matrix factorization, compressive sensing, inverse problems, blind source separation}

@@

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Coulter10, Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if all dictionaries consistent with training data do not differ appreciably in their ability to sparsely encode new samples. 
Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain this uniqueness phenomenon more generally.

@@

Note that any particular solution $(\mathbf{B}, \mathbf{\bar x}_1 \ldots, \mathbf{\bar x}_N)$ to this problem gives rise to an orbit of equivalent solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal. Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) has shown that a solution to Prob.~\ref{InverseProblem} (when it exists) is indeed unique up to this inherent ambiguity provided the $\mathbf{x}_i$ are sufficiently diverse and the generating matrix $\mathbf{A}$ satisfies the \textit{spark condition} from compressive sensing:
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
which would be, in any case, necessary for uniqueness of the $\mathbf{x}_i$. Our concern here is solution stability with respect to noise.

@@

We generalize this domain-restricted lower bound to a \emph{union of subspaces} model \cite{vidal2005generalized} derived from the hypergraph $\mathcal{H}$. Letting $\mathbf{M}_J$ denote the submatrix formed by the columns of $\mathbf{M}$ indexed by $J$ (with $\mathbf{M}_\emptyset := \mathbf{0}$), we define: 
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressive sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \alpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \frac{1}{\sqrt{k}} \inf \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
\end{align} 
%
where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressive sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$. Note that if $\mathcal{H}$ covers $[m]$ then $L_2 \geq L_{2\mathcal{H}}$.

@@

Every real $n \times \bar m$ matrix $\mathbf{B}$ for which there exist $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N$ satisfying \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$} for all $i$ has $\bar m \geq m$ and, provided $\bar m < mr/(r-1)$,
\begin{align}\label{Cstable}
\|\mathbf{A}_j- \mathbf{B}_{\bar J} \mathbf{PD}_j\|_2 \leq C_1 \varepsilon \ \ \text{for all } j \in J,
\end{align}
%
for some non-empty $J, \bar J$ of size $\bar m - r(\bar m - m)$, some permutation $\mathbf{P}$ and invertible diagonal $\mathbf{D}$.

@@

%Regarding the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, 
It so happens that producing sparse codes $\mathbf{x}_i$ in general linear position is straightforward with a ``Vandermonde'' matrix construction (e.g., see \cite{Hillar15}). We therefore have:

@@

%%%%%%%%%%%%%%
% need to cut maybe: 
%As mentioned in the previous section, Thm.~\ref{DeterministicUniquenessTheorem} is a particular case of a more general result that forgoes the assumption that $\mathcal{H} \subseteq {[m] \choose k}$ is regular and satisfies the SIP. If instead we require only that the stars $\cap h(i)$ intersect at singletons for all $i \leq q$ (assuming that the nodes of $\mathcal{H}$ are labeled in some order of non-increasing degree), we have that $\bar m \geq k|\mathcal{H}| / \deg(1)$ and, provided $\bar m < k|\mathcal{H}| / (\deg(1) - 1)$, the non-empty submatrix $J$ is of size equal to the largest number $p$ satisfying:
%\begin{align}\label{pcond}
%\sum_{i=\ell}^{m} \deg(i) > (\bar m + 1 - \ell) (\deg(\ell) - 1) \ \ \text{for all } \ell \leq p \leq q.
%\end{align}
%Specifically, $J$ contains all nodes of degree exceeding $\deg(p)$ and some subset of those with degreee equal to $\deg(p)$. For the benefit of the reader, we do not prove explicitly this more general result below; it can be discerned from how exactly Lemma \ref{NonEmptyLemma} is incorporated into the proof of Lem.~\ref{MainLemma}.
%%%%%%%%%%%%%%%%

@@

If for some  $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and map $\pi: E \mapsto {[\bar m] \choose k}$ we have:
\begin{align}\label{GapUpperBound}
d(\bm{\mathcal{A}}_S, \bm{\mathcal{B}}_{\pi(S)}) \leq \varepsilon, \ \  \text{for $S \in \mathcal{H}$},
\end{align}
then $\bar m \geq m$, and provided $\bar m < mr /(r-1)$, there exists a permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that:
\begin{align}\label{MainLemmaBPD}
\|\mathbf{A}_j - \mathbf{B}_{\bar J} \mathbf{PD}_j\|_2 \leq C_2 \varepsilon, \ \  \text{for } j \in J,
\end{align}
for some non-empty $J, \bar J$ of size $\bar m - r(\bar m - m)$.
\end{lemma}

@@

Since there are more than $(k-1){\bar m \choose k}$ vectors $\mathbf{x}_i$ supported in $S$, by the pigeonhole principle there must be some $\bar S \in {[\bar m] \choose k}$ and set of $k$ indices $K \subseteq I(S)$ such that all $\mathbf{\bar x}_i$ with $i \in K$ are supported in $\bar S$.

It follows from $L_{2\mathcal{H}}(\mathbf{A}) > 0$ and the general linear position of the $\mathbf{x}_i$ that $L(\mathbf{AX}_{K}) > 0$; that is, the columns of the $n \times k$ matrix $\mathbf{AX}_K$ form a basis for $\bm{\mathcal{A}}_S$. Fixing $\mathbf{0} \neq \mathbf{y} \in \bm{\mathcal{A}}_S$, there then exists $\mathbf{0} \neq \mathbf{c} = (c_1, \ldots, c_k) \in \mathbb{R}^k$ such that $\mathbf{y} = \mathbf{AX}_K\mathbf{c}$. Setting \mbox{$\mathbf{\bar{y}} = \mathbf{B\bar{X}}_K\mathbf{c} \in \bm{\mathcal{B}}_{\bar S}$}, we have:
\begin{align*}
\|\mathbf{y} - \mathbf{\bar{y}}\|_2 
&= \|\sum_{i=1}^k c_i(\mathbf{AX}_K - \mathbf{B\bar{X}}_K)_i\|_2
\leq \varepsilon \sum_{i=1}^k |c_i| \\
&\leq \varepsilon \sqrt{k}  \|\mathbf{c}\|_2 
\leq \frac{\varepsilon}{L(\mathbf{AX}_K)} \|\mathbf{y}\|_2,
\end{align*}
where the last inequality follows directly from \eqref{Ldef}. From Def.~\ref{dDef} we have:
\begin{align}\label{rhs222}
d(\bm{\mathcal{A}}_S, \bm{\mathcal{B}}_{\bar S}) 
\leq \frac{\varepsilon}{  L(\mathbf{AX}_{K}) } \leq \varepsilon \frac{C_1}{C_2},
\end{align}
%
where the second inequality is due to $L(\mathbf{AX}_{K}) \geq L_k(\mathbf{AX}_{I(S)})$ and \eqref{Cdef1}. Since $\varepsilon < L_2(\mathbf{A})/C_1$, the result follows by Lem.~\ref{MainLemma}, .
\end{proof}

@@

In this note, we prove the stability of unique solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} while significantly reducing the known sample complexity. Our results justify the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research. The main motivation for this work, however, was to understand how seemingly universal representations emerge from sparse coding models fit to natural data by a variety of methodologies. We elaborate on these applications below after discussing theoretical aspects.

What we have shown here is that the sparse linear coding model generally produces a \textit{well-posed} inverse problem to be approximately solved by a numerical algorithm. This early concept of Hadamard \cite{Hadamard1902} can be paraphrased as the idea that inferences from observations should be robust to the inevitable uncertainty in measurement. In other words, a small perturbation of the data should result in only slightly different inferred parameters. In this regard, for the sparse linear coding model we demonstrate a linear relationship, \eqref{epsdel}, between measurement noise and the deviation of any solution from the true parameters, with explicit constants expressed in terms of these parameters. Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise. It would therefore be of practical utility to determine the best possible dependence of $\varepsilon$ on $\delta_1, \delta_2$ in Def.\ref{maindef} as well as the minimal requirements on the number and diversity of generating codes, and we hope that other researchers continue to improve and extend our results. We remark that our constants have been derived for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be smaller when sampled from a distribution; in such cases, the constants will improve as well.

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singleton intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. For instance, whether there is a sample size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}. A re-examination of pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this can in fact be achieved if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have applied compressive sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressive sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems.

In the field of theoretical neuroscience in particular, dictionary learning for sparse coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that reproduce response properties of cortical neurons. Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early mathematical idea in neural theory \cite{pitts1947}. Furthermore, they justify the hypothesis of \cite{Coulter10, Isely10} that sparse codes passed through information bottlenecks in the brain are recovered from random projections via (unsupervised) biologically plausible sparse coding (e.g., \cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}).

Deep learning noise issues: \cite{szegedy2013intriguing, goodfellow2014explaining}.

@@

[*** TODO: make journals abbreviated references ***]

@@

If the map $\pi: E \to 2^{[\bar m]}$ has $\sum_{S \in \mathcal{H}} (|\pi(S)| -|S|) \geq 0$ and:
\begin{align}\label{cond}
|\cap \pi(\mathcal{G})| \leq |\cap \mathcal{G} | \text{ for all } \mathcal{G} \subseteq \mathcal{H},
\end{align}
then $\bar m \geq\sum_i \deg(i) / \deg(1)$, and if $\bar m < \sum_i \deg(i) / (\deg(1) - 1)$ then the association $i \mapsto \cap \pi(h(i))$ defines an injective map to $[\bar m]$ from some non-empty $J \subseteq [m]$ of size $p$ consisting of the union of the set of all nodes of degree exceeding $\deg(p)$ and some set of nodes all having degree equal to $\deg(p)$. In particular, if $\mathcal{H}$ is $r$-regular then $|J| = \bar m - r(\bar m - m)$. 
\end{lemma}

@@

By \eqref{pcond} and the upper bound on $\bar m$ we have $|T_1| >  \bar m (\deg(1) - 1)$, which implies, again by the pigeonhole principle, that there must be at least $\deg(1)$ elements of $T_1$ sharing the same first index. By \eqref{cond}, the intersection of the set $\mathcal{G}_1$ consisting of their second indices is non-empty. As $p \leq q$ and $\deg(1) \geq \deg(i)$ for all $i$, it must be that $|\cap \mathcal{G}_1| = 1$. Since $\cap \pi(\mathcal{G}_1)$ is non-empty, applying \eqref{cond} again implies $\cap \pi(\mathcal{G}_1) = \{i_1\}$ for some $i_1 \in [m]$. If $p=1$ then we are done. Otherwise, define $T_2 := T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains $|T_2| = |T_1| - \deg(1) = \sum_{i=2}^m \deg(i)$ ordered pairs having $\bar m - 1$ distinct first indices. By \eqref{pcond} we have $|T_2| > (\bar m - 1)(\deg(2) - 1)$ and reiterating the above arguments produces a (necessarily) distinct index $i_2$. Iterating the arguments $p$ times yields the set of singletons \mbox{$J = \{\cap \mathcal{G}_1, \ldots, \cap \mathcal{G}_p\} \subseteq [m]$}.
%TODO: Let T_1 be the set of all pairs $T_1 := \{(R, S): R \subseteq \pi(S), S \in \mathcal{H}\}$ 
\end{proof}

>>>>

\keywords{Sparse coding, dictionary learning, matrix factorization, compressed sensing, inverse problems, blind source separation}

@@

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if all dictionaries consistent with training data do not differ appreciably in their ability to sparsely encode new samples. 
Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain this uniqueness phenomenon more generally.

@@

Note that any particular solution $(\mathbf{B}, \mathbf{\bar x}_1 \ldots, \mathbf{\bar x}_N)$ to this problem gives rise to an orbit of equivalent solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal. Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) has shown that a solution to Prob.~\ref{InverseProblem} (when it exists) is indeed unique up to this inherent ambiguity provided the $\mathbf{x}_i$ are sufficiently diverse and the generating matrix $\mathbf{A}$ satisfies the \textit{spark condition} from compressed sensing:
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
which would be, in any case, necessary for uniqueness of the $\mathbf{x}_i$. Our concern here is solution stability with respect to noise.

@@

We generalize this domain-restricted lower bound to a \emph{union of subspaces} model \cite{vidal2005generalized} derived from the hypergraph $\mathcal{H}$. Letting $\mathbf{M}_J$ denote the submatrix formed by the columns of $\mathbf{M}$ indexed by $J$ (with $\mathbf{M}_\emptyset := \mathbf{0}$), we define: 
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \inf \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \alpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \frac{1}{\sqrt{k}} \inf \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
\end{align} 
%
where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$. Note that if $\mathcal{H}$ covers $[m]$ then $L_2 \geq L_{2\mathcal{H}}$.

@@

Every real $n \times \bar m$ matrix $\mathbf{B}$ for which there exist $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N$ satisfying \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$} for all $i$ has $\bar m \geq m$ and, provided $\bar m < mr/(r-1)$,
\begin{align}\label{Cstable}
\|\mathbf{A}_j- \mathbf{B}_{\bar J} \mathbf{PD}_j\|_2 \leq C_1 \varepsilon \ \ \text{for all } j \in J,
\end{align}
%
for some nonempty $J, \bar J$ of size $\bar m - r(\bar m - m)$, some permutation $\mathbf{P}$ and invertible diagonal $\mathbf{D}$.

@@

%Regarding the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, 
It so happens that producing sparse codes $\mathbf{x}_i$ in general linear position is straightforward with a ``Vandermonde'' matrix construction (e.g., see \cite{Hillar15}).

@@

%%%%%%%%%%%%%%
% need to cut maybe: 
%As mentioned in the previous section, Thm.~\ref{DeterministicUniquenessTheorem} is a particular case of a more general result that forgoes the assumption that $\mathcal{H} \subseteq {[m] \choose k}$ is regular and satisfies the SIP. If instead we require only that the stars $\cap h(i)$ intersect at singletons for all $i \leq q$ (assuming that the nodes of $\mathcal{H}$ are labeled in some order of non-increasing degree), we have that $\bar m \geq k|\mathcal{H}| / \deg(1)$ and, provided $\bar m < k|\mathcal{H}| / (\deg(1) - 1)$, the nonempty submatrix $J$ is of size equal to the largest number $p$ satisfying:
%\begin{align}\label{pcond}
%\sum_{i=\ell}^{m} \deg(i) > (\bar m + 1 - \ell) (\deg(\ell) - 1) \ \ \text{for all } \ell \leq p \leq q.
%\end{align}
%Specifically, $J$ contains all nodes of degree exceeding $\deg(p)$ and some subset of those with degreee equal to $\deg(p)$. For the benefit of the reader, we do not prove explicitly this more general result below; it can be discerned from how exactly Lemma \ref{NonEmptyLemma} is incorporated into the proof of Lem.~\ref{MainLemma}.
%%%%%%%%%%%%%%%%

@@

If for some  $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and map $\pi: E \mapsto {[\bar m] \choose k}$ we have:
\begin{align}\label{GapUpperBound}
d(\bm{\mathcal{A}}_S, \bm{\mathcal{B}}_{\pi(S)}) \leq \varepsilon, \ \  \text{for $S \in \mathcal{H}$},
\end{align}
then $\bar m \geq m$, and provided $\bar m < mr /(r-1)$, there is a permutation matrix $\mathbf{P}$ and invertible diagonal $\mathbf{D}$ such that:
\begin{align}\label{MainLemmaBPD}
\|\mathbf{A}_j - \mathbf{B}_{\bar J} \mathbf{PD}_j\|_2 \leq C_2 \varepsilon, \ \  \text{for } j \in J,
\end{align}
for some nonempty $J, \bar J$ of size $\bar m - r(\bar m - m)$.
\end{lemma}

@@

Since there are more than $(k-1){\bar m \choose k}$ vectors $\mathbf{x}_i$ supported in $S$, the pigeonhole principle gives $\bar S \in {[\bar m] \choose k}$ and a set of $k$ indices $K \subseteq I(S)$ with all $\mathbf{\bar x}_i$, $i \in K$, supported in $\bar S$.

It follows from $L_{2\mathcal{H}}(\mathbf{A}) > 0$ and the general linear position of the $\mathbf{x}_i$ that $L(\mathbf{AX}_{K}) > 0$; that is, the columns of the $n \times k$ matrix $\mathbf{AX}_K$ form a basis for $\bm{\mathcal{A}}_S$. Fixing $\mathbf{0} \neq \mathbf{y} \in \bm{\mathcal{A}}_S$, there then exists $\mathbf{0} \neq \mathbf{c} = (c_1, \ldots, c_k) \in \mathbb{R}^k$ such that $\mathbf{y} = \mathbf{AX}_K\mathbf{c}$. Setting \mbox{$\mathbf{\bar{y}} = \mathbf{B\bar{X}}_K\mathbf{c} \in \bm{\mathcal{B}}_{\bar S}$}, we have:
\begin{align*}
\|\mathbf{y} - \mathbf{\bar{y}}\|_2 
&= \|\sum_{i=1}^k c_i(\mathbf{AX}_K - \mathbf{B\bar{X}}_K)_i\|_2
\leq \varepsilon \sum_{i=1}^k |c_i| \\
&\leq \varepsilon \sqrt{k}  \|\mathbf{c}\|_2 
\leq \frac{\varepsilon}{L(\mathbf{AX}_K)} \|\mathbf{y}\|_2,
\end{align*}
where the last inequality follows directly from \eqref{Ldef}. From Def.~\ref{dDef} we have:
\begin{align}\label{rhs222}
d(\bm{\mathcal{A}}_S, \bm{\mathcal{B}}_{\bar S}) 
\leq \frac{\varepsilon}{  L(\mathbf{AX}_{K}) } \leq \varepsilon \frac{C_1}{C_2},
\end{align}
%
where the second inequality is due to $L(\mathbf{AX}_{K}) \geq L_k(\mathbf{AX}_{I(S)})$ and \eqref{Cdef1}. Finally, apply Lem.~\ref{MainLemma} with $\varepsilon < L_2(\mathbf{A})/C_1$.
\end{proof}

@@

In this note, we prove the stability of unique solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} while significantly reducing the known sample complexity. Our results justify the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research. The main motivation for this work, however, was to understand how seemingly universal representations emerge from sparse coding models fit to natural data by a variety of methodologies. We elaborate on these applications below.

What we have shown here is that the sparse linear coding model generally produces a \textit{well-posed} inverse problem to be approximately solved by a numerical algorithm. This early concept of Hadamard \cite{Hadamard1902} can be paraphrased as the idea that inferences from observations should be robust to the inevitable uncertainty in measurement. In other words, a small perturbation of the data should result in only slightly different inferred parameters. In this regard, for the sparse linear coding model we demonstrate a linear relationship, \eqref{epsdel}, between measurement noise and the deviation of any solution from the true parameters, with explicit constants expressed in terms of these parameters. Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise. It would therefore be of practical utility to determine the best possible dependence of $\varepsilon$ on $\delta_1, \delta_2$ in Def.\ref{maindef} as well as the minimal requirements on the number and diversity of generating codes, and we hope that other researchers continue to improve and extend our results. We remark that our constants have been derived for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be smaller when sampled from a distribution.  % ; in such cases, the constants will improve as well.

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singleton intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. For instance, whether there is a sample size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}. A reexamination of pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this can in fact be achieved if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems.

In the field of theoretical neuroscience in particular, dictionary learning for sparse coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12} that reproduce response properties of cortical neurons. Our results suggest that this correspondence could be due to the ``universality'' of sparse representations in natural data, an early mathematical idea in neural theory \cite{pitts1947}. Furthermore, they justify the hypothesis of \cite{Isely10} that sparse codes passed through information bottlenecks in the brain are recovered from random projections via (unsupervised) biologically plausible sparse coding (e.g., \cite{rehnsommer2007, rozell2007neurally, ganguli2012compressed, hu2014hebbian}).

Deep learning noise issues: \cite{goodfellow2014explaining}.

@@

@@

If the map $\pi: E \to 2^{[\bar m]}$ has $\sum_{S \in \mathcal{H}} (|\pi(S)| -|S|) \geq 0$ and:
\begin{align}\label{cond}
|\cap \pi(\mathcal{G})| \leq |\cap \mathcal{G} | \text{ for all } \mathcal{G} \subseteq \mathcal{H},
\end{align}
then $\bar m \geq\sum_i \deg(i) / \deg(1)$, and if $\bar m < \sum_i \deg(i) / (\deg(1) - 1)$ then the association $i \mapsto \cap \pi(h(i))$ defines an injective map to $[\bar m]$ from some nonempty $J \subseteq [m]$ of size $p$ consisting of the union of the set of all nodes of degree exceeding $\deg(p)$ and some set of nodes all having degree equal to $\deg(p)$. In particular, if $\mathcal{H}$ is $r$-regular then $|J| = \bar m - r(\bar m - m)$. 
\end{lemma}

@@

By \eqref{pcond} and the upper bound on $\bar m$ we have $|T_1| >  \bar m (\deg(1) - 1)$, which implies, again by the pigeonhole principle, that there must be at least $\deg(1)$ elements of $T_1$ sharing the same first index. By \eqref{cond}, the intersection of the set $\mathcal{G}_1$ consisting of their second indices is nonempty. As $p \leq q$ and $\deg(1) \geq \deg(i)$ for all $i$, it must be that $|\cap \mathcal{G}_1| = 1$. Since $\cap \pi(\mathcal{G}_1)$ is nonempty, applying \eqref{cond} again implies $\cap \pi(\mathcal{G}_1) = \{i_1\}$ for some $i_1 \in [m]$. If $p=1$ then we are done. Otherwise, define $T_2 := T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains $|T_2| = |T_1| - \deg(1) = \sum_{i=2}^m \deg(i)$ ordered pairs having $\bar m - 1$ distinct first indices. By \eqref{pcond} we have $|T_2| > (\bar m - 1)(\deg(2) - 1)$ and reiterating the above arguments produces a (necessarily) distinct index $i_2$. Iterating the arguments $p$ times yields the set of singletons \mbox{$J = \{\cap \mathcal{G}_1, \ldots, \cap \mathcal{G}_p\} \subseteq [m]$}.
%TODO: Let T_1 be the set of all pairs $T_1 := \{(R, S): R \subseteq \pi(S), S \in \mathcal{H}\}$ 
\end{proof}
